---
title: "TSDataProject"
output: html_document
author: "Jacob Croskey"
date: "2024-05-06"
---

```{r setup, include=FALSE}
setwd("~//Desktop/Time Series")
library(dplyr)
library(tidyverse)
library(lubridate)
library("readxl")
library("tseries")
library("astsa")
library(forecast)
library(vars)
library(urca)
library(dynamac)
library(TSstudio)
library(stargazer)
library(ARDL)
library(pbapply)
library(AICcmodavg)
library(ggplot2)
library(MASS)
```

## Reading in Data
### I cleaned some of the data in this chunk as well to prepare them to be transformed into Time Series.
```{r}
#Average Temp Data and Transformation
avgtemp <- read.csv("tempdata.csv")
avgtemp$Date <- as.Date(paste0(avgtemp$Date, "-01"))
avgtemp <- avgtemp %>% 
  rename(temp = Value)

#Average Ozone Data and Transformation
airqual <- read.csv("pollution_2000_2023.csv")
ozone <- airqual[c("State", "Date", "O3.Mean")]
ozone$Date <- as.Date(ozone$Date)
monthlyozone <- ozone %>%
  filter(day(Date) == 1)
amo <- monthlyozone %>%
  group_by(Date) %>%
  summarise(avgO3 = mean(O3.Mean, na.rm = TRUE), .groups = 'drop')
```

## Plots of Temperatures and Ozone Levels
```{r}
plot(amo)
plot(avgtemp)
```

## Transforming Data Into Time Series And Checking for Stationarity
```{r}
tempf = ts(data = avgtemp$temp, frequency = 12, start = c(2000,1), end = c(2023, 1))
plot(tempf)
lines(lowess(tempf), col = "red")

ozonelevel = ts(data = amo$avgO3, frequency = 12, start = c(2000,1), end = c(2023, 1))
plot(ozonelevel)
lines(lowess(ozonelevel), col = "blue")

pp.test(tempf)
pp.test(ozonelevel)
```

## Transformation and Combination of Data into a Time Series for Analyzation
```{r}
variables <- ts.union(tempf, ozonelevel)
plot(variables)
start1 = start = c(2000,1); end1 = end = c(2023,1)

variables_df <- data.frame(date = time(variables),
                           Temp = as.numeric(variables[, "tempf"]),
                           Ozone = as.numeric(variables[, "ozonelevel"]))

# Convert 'date' from fractional years to actual dates (assuming monthly data starting January 2000)
variables_df$date <- as.Date(ISOdate(2000, 1, 1) + (variables_df$date - 2000) * 365.25)
variables_df$month <- format(variables_df$date, "%m")

# Create dummy variables for months
month_dummies <- model.matrix(~ factor(month) - 1, data = variables_df)
variables_df <- cbind(variables_df, month_dummies)
```

## Creating Lagged Variables and Releveling NA Values Introduced by Lagging
```{r}
variables_df$lag_Ozone = lag(variables_df$Ozone, 1)
variables_df$lag_Temp = lag(variables_df$Temp, 1)
variables_df <- na.omit(variables_df)

#Setting Month as Factor and Setting January for Reference Month to eliminate Perfect Colinearity
variables_df$month <- as.factor(variables_df$month)
variables_df$month <- relevel(variables_df$month, ref = "01")
```


```{r eval=FALSE, include=FALSE}
## Checking Optimal Number of Lags using BIC/AIC Criterion
#Checking Optimal Number of Lags with BIC
models <- list()
bic_values <- numeric()

for (i in 1:6) {  # testing up to 6 lags
  formula <- as.formula(
    paste("Ozone ~ lag(Ozone, 1:", i, ") + Temp + lag(Temp, 1:", i, ") + factor(month)", collapse = "+")
  )
  models[[i]] <- lm(formula, data = variables_df)
  bic_values[i] <- BIC(models[[i]])
}
# Which model has the lowest BIC?
best_lag <- which.min(bic_values)
best_lag 

#Checking Optimal Number of Lags with AIC
#install.packages("pbapply")
library(pbapply)
#install.packages("AICcmodavg")
library(AICcmodavg)
models <- list()
aic_values <- numeric()

for (i in 1:6) {  # testing up to 6 lags
  formula <- as.formula(
    paste("Ozone ~ lag(Ozone, 1:", i, ") + lag(Temp, 1:", i, ") + factor(month) - 1", collapse = "+")
  )
  models[[i]] <- lm(formula, data = variables_df)
  aic_values[i] <- AICc(models[[i]])
}
# Which model has the lowest AIC?
which.min(aic_values)
```

## Specifying ARDL Model with Lags and Dummies 
```{r}
ardl_model <- lm(Ozone ~ lag_Ozone + Temp + lag_Temp + factor(month), data = variables_df)

summary(ardl_model)
plot(ardl_model)
```

## AR Forecast of Temperature for 12 Months Out
```{r}
# Fit AR model automatically selecting optimal lags
ar_model_xt <- auto.arima(variables_df$Temp, seasonal = TRUE)

summary(ar_model_xt)

forecast_xt <- forecast(ar_model_xt, h = 12, level = 95)

# Plot the forecasts including 95% prediction intervals
plot(forecast_xt, main = "Temperature Forecast with 95% Prediction Interval", xlab = "Time (Months)", ylab = "Temperature (°F)")
print(forecast_xt)
```

## AR Forecast of Ozone Levels
```{r}
# Build the linear model using Ozone as dependent variable and Temperature, lagged Ozone, and month dummies as predictors
linear_model_yt <- lm(Ozone ~ Temp + lag_Ozone + factor(month) - 1, data = variables_df)

# View the summary of the model to check for significance and fit
summary(linear_model_yt)

# Assuming forecast_xt contains forecasted temperatures and forecast_dates are the corresponding future dates
forecast_data <- data.frame(
  Temp = forecast_xt$mean,  # Forecasted temperatures from AR model
  lag_Ozone = rep(tail(variables_df$Ozone, 1), 12),  # Repeat last known ozone value
  date = seq.Date(from = tail(variables_df$date, 1) + 1, by = "month", length.out = 12)
)

# Convert dates to month factors if your model includes monthly effects
forecast_data$month <- factor(format(forecast_data$date, "%m"), levels = levels(variables_df$month))

# Recreate the dummy variables for the months (if your model includes this)
month_dummies <- model.matrix(~ month - 1, data = forecast_data)
forecast_data <- cbind(forecast_data, month_dummies[, -1])  # Exclude intercept column

yt_forecasts <- predict(linear_model_yt, newdata = forecast_data, interval = "prediction")

# Create a data frame to store the results for better visualization
results_df <- data.frame(
  Date = forecast_data$date,
  Ozone_Forecast = yt_forecasts[, "fit"],
  Lower_95_CI = yt_forecasts[, "lwr"],
  Upper_95_CI = yt_forecasts[, "upr"]
)

# Print the forecast results
print(results_df)

#Plotting Forecast
ggplot(results_df, aes(x = Date+30)) +
  geom_line(aes(y = Ozone_Forecast), color = "blue") +
  geom_ribbon(aes(ymin = Lower_95_CI, ymax = Upper_95_CI), fill = "blue", alpha = 0.3) +
  labs(title = "Forecasted Ozone Levels", x = "Date", y = "Ozone Level")
```

